---
title: "Analysis"
output: html_document
---

```{r}
library(tidyverse)
library(readr)
collegescorecard <- read_csv("group-h-csv.csv", na = c("", "NA", "NULL", "PrivacySuppressed"),
    col_types = cols(ADM_RATE = col_number(), 
        ADM_RATE_ALL = col_number(), C150_4 = col_number(), 
        COMPL_RPY_3YR_RT_SUPP = col_number(), 
        CONTROL = col_character(), COSTT4_A = col_number(), 
        DEP_RPY_3YR_RT_SUPP = col_number(), 
        FAMINC = col_number(), FEMALE_RPY_3YR_RT_SUPP = col_number(), 
        FIRSTGEN_RPY_3YR_RT_SUPP = col_number(), 
        GRAD_DEBT_MDN10YR_SUPP = col_number(), 
        GRAD_DEBT_MDN_SUPP = col_number(), 
        HIGHDEG = col_character(), HI_INC_RPY_3YR_RT_SUPP = col_number(), 
        ICLEVEL = col_character(), IND_RPY_3YR_RT_SUPP = col_number(), 
        LO_INC_RPY_3YR_RT_SUPP = col_number(), 
        MALE_RPY_3YR_RT_SUPP = col_number(), 
        MD_EARN_WNE_P10 = col_number(), MD_FAMINC = col_number(), 
        MD_INC_RPY_3YR_RT_SUPP = col_number(), 
        MN_EARN_WNE_P10 = col_number(), NONCOM_RPY_3YR_RT_SUPP = col_number(), 
        NOPELL_RPY_3YR_RT_SUPP = col_number(), 
        NOTFIRSTGEN_RPY_3YR_RT_SUPP = col_number(), 
        NPT4_PRIV = col_number(), NPT4_PUB = col_number(), 
        OPEID = col_character(), PCTFLOAN = col_number(), 
        PCTPELL = col_number(), PELL_RPY_3YR_RT_SUPP = col_number(), 
        PFTFAC = col_number(), PREDDEG = col_character(), 
        RPY_3YR_RT_SUPP = col_number(), SAT_AVG = col_number(), 
        TUITFTE = col_number(), TUITIONFEE_IN = col_number(), 
        TUITIONFEE_OUT = col_number(), UNITID = col_character()))

collegescorecard<- collegescorecard%>% mutate(SCH_DEG=as.factor(SCH_DEG), PREDDEG=as.factor(PREDDEG), HIGHDEG=as.factor(HIGHDEG), CONTROL=as.factor(CONTROL)) 

```

```{r}
library(ggplot2)
library(ggmap)
library(maps)

statedata<-collegescorecard%>%
  group_by(STABBR)%>%
  summarise(total=mean(MN_EARN_WNE_P10, na.rm=TRUE))


ggplot(data=statedata, aes(x=STABBR, y=total))+
  geom_point()

```


```{r}
model<-lm(MN_EARN_WNE_P10~MD_FAMINC+PCTPELL+GRAD_DEBT_MDN10YR_SUPP+ADM_RATE+LO_INC_RPY_3YR_RT_SUPP, data=collegescorecard)

repaymentmod<-lm(MN_EARN_WNE_P10~MALE_RPY_3YR_RT_SUPP+FEMALE_RPY_3YR_RT_SUPP+FIRSTGEN_RPY_3YR_RT_SUPP+HI_INC_RPY_3YR_RT_SUPP+LO_INC_RPY_3YR_RT_SUPP+IND_RPY_3YR_RT_SUPP+NONCOM_RPY_3YR_RT_SUPP, data=collegescorecard)

repay<-lm(MN_EARN_WNE_P10~HI_INC_RPY_3YR_RT_SUPP+LO_INC_RPY_3YR_RT_SUPP+IND_RPY_3YR_RT_SUPP, data=collegescorecard)

anova(repay)

anova(repaymentmod)

plot(repay)

summary(repaymentmod)
summary(model)

library(ggplot2)
ggplot(data=collegescorecard, aes(x=FAMINC))+
  geom_point(aes(y=HI_INC_RPY_3YR_RT_SUPP, color="HI_INC_RPY_3YR_RT_SUPP"))+
  geom_point(aes(y=LO_INC_RPY_3YR_RT_SUPP, color="LO_INC_RPY_3YR_RT_SUPP"))+
  geom_point(aes(y=IND_RPY_3YR_RT_SUPP, color="IND_RPY_3YR_RT_SUPP"))
```

## Explanatory Variables: 

Mean family income: quantitative Unit: dollars
Using data from NSLDS, these elements calculate the percentage of Title IV-receiving students who had family income data from the Free Application for Federal Student Aid (FAFSA) for each of five income quintiles.

Tuition cost: quantitative Unit: dollars
Includes tuition and fees, books, and supplies, and living expenses for all full-time, first-time, degree/certificate-seeking undergraduates who receive Title IV aid.

Percent of students on Pell grants: quantitative Unit: percent
Calculated from values from the IPEDS Student Financial Aid Component.

Percent of students receiving federal loan: quantitative Unit: percent
Calculated from values from the IPEDS Student Financial Aid Component

Cumulative mean debt: quantitative Unit: dollars 
Represents the sum of all undergraduate federal loans over students’ college education at the institution for which the median debt is reported for —e.g., if a student receives a federal loan for $2,000 for each of eight semesters at one institution, their cumulative debt is recorded as 
$16,000 for that institution. 

School Type: categorical Categories: private, public, trade



https://www.aauw.org/research/deeper-in-debt/ 


=======
### Model Attempt: FamInc


```{r}
modcor<-lm(TUITIONFEE_IN~TUITIONFEE_OUT, data=collegescorecard)
summary(modcor)
```

Out of state tuition was removed from the model because it was highly multicolinear with in state tuition. No other significant multicolinarity ($R^2>0.8$) was found.

```{r}
mod1<- lm(MN_EARN_WNE_P10~MD_FAMINC+SCH_DEG+ADM_RATE_ALL+C150_4+HIGHDEG+TUITFTE+TUITIONFEE_IN+SAT_AVG+GRAD_DEBT_MDN_SUPP+COSTT4_A+PFTFAC+PCTPELL+NPT4_PUB+PCTFLOAN+RPY_3YR_RT_SUPP, data=collegescorecard)
summary(mod1)
```

C150_4 was removed because it had the largest p-value.

```{r}
mod2<- lm(MN_EARN_WNE_P10~MD_FAMINC+SCH_DEG+ADM_RATE_ALL+HIGHDEG+TUITFTE+TUITIONFEE_IN+SAT_AVG+GRAD_DEBT_MDN_SUPP+COSTT4_A+PFTFAC+PCTPELL+NPT4_PUB+PCTFLOAN+RPY_3YR_RT_SUPP, data=collegescorecard)
summary(mod2)
```
Median Grad debt was removed from the model. p=0.81865

```{r}
mod3<- lm(MN_EARN_WNE_P10~MD_FAMINC+SCH_DEG+ADM_RATE_ALL+HIGHDEG+TUITFTE+TUITIONFEE_IN+SAT_AVG+COSTT4_A+PFTFAC+PCTPELL+NPT4_PUB+PCTFLOAN+RPY_3YR_RT_SUPP, data=collegescorecard)
summary(mod3)
```

Most common degree awarded was removed from the model.

```{r}
mod4<- lm(MN_EARN_WNE_P10~MD_FAMINC+ADM_RATE_ALL+HIGHDEG+TUITFTE+TUITIONFEE_IN+SAT_AVG+COSTT4_A+PFTFAC+PCTPELL+NPT4_PUB+PCTFLOAN+RPY_3YR_RT_SUPP, data=collegescorecard)
summary(mod4)
```
In state tuition was removed.

```{r}
mod5<- lm(MN_EARN_WNE_P10~MD_FAMINC+ADM_RATE_ALL+HIGHDEG+TUITFTE+SAT_AVG+COSTT4_A+PFTFAC+PCTPELL+NPT4_PUB+PCTFLOAN+RPY_3YR_RT_SUPP, data=collegescorecard)
summary(mod5)
```

PFTFAC was removed

```{r}
mod6<- lm(MN_EARN_WNE_P10~MD_FAMINC+ADM_RATE_ALL+HIGHDEG+TUITFTE+SAT_AVG+COSTT4_A+PCTPELL+NPT4_PUB+PCTFLOAN+RPY_3YR_RT_SUPP, data=collegescorecard)
summary(mod6)
```
NPT4_PUB

```{r}
mod7<- lm(MN_EARN_WNE_P10~MD_FAMINC+ADM_RATE_ALL+HIGHDEG+TUITFTE+SAT_AVG+COSTT4_A+PCTPELL+PCTFLOAN+RPY_3YR_RT_SUPP, data=collegescorecard)
summary(mod7)
```
COST4_A


```{r}
mod8<- lm(MN_EARN_WNE_P10~MD_FAMINC+ADM_RATE_ALL+HIGHDEG+TUITFTE+SAT_AVG+PCTPELL+PCTFLOAN+RPY_3YR_RT_SUPP, data=collegescorecard)
summary(mod8)
```

Highdeg

```{r}
mod9<- lm(MN_EARN_WNE_P10~MD_FAMINC+ADM_RATE_ALL+TUITFTE+SAT_AVG+PCTPELL+PCTFLOAN+RPY_3YR_RT_SUPP, data=collegescorecard)
summary(mod9)
```



Based on Model 9, median family income, admit rate, tuition revenue per full time student, average SAT score, percent of students receiving pell grants, percent of students receiving federal loans, and 3 year repayment rate are all significantly associated with mean earnings 10 years after graduation.

## Assessing Conditions for Regression:

```{r}
qqnorm(residuals(mod9))
  qqline(residuals(mod9))
plot(residuals(mod9)~fitted(mod9))
  abline(h=0)
  hist(residuals(mod9))
```

** Warning: This code takes a long time to run!:** If you want to knit without running it, put "{r, eval=FALSE}" where it normally says "{r}".

##Correlation Matrix

```{r, eval=FALSE}
library(GGally)
ggpairs(mod1)
```

I used a GGPairs plot to look for multicolinearity in our model. The plot was difficult to visualize because we had so many variables in the initial model, but I think that it's important to start by looking at all of the variables if we are using backwards selection based on p-values. P-values are less reliable when multicolinearity is suspected. 

I tried to visualize the plot by downloading it as a png file, but that didn't make it easier to read. 

```{r, eval=FALSE}
library(GGally)
ggpairs(mod7)
plot(mod7)
```


## Attempting backward selection:
```{r}
library(leaps)
full <- lm(MN_EARN_WNE_P10~MD_FAMINC+SCH_DEG+ADM_RATE_ALL+C150_4+HIGHDEG+TUITFTE+TUITIONFEE_IN+SAT_AVG+GRAD_DEBT_MDN_SUPP+COSTT4_A+PFTFAC+PCTPELL+NPT4_PUB+PCTFLOAN+RPY_3YR_RT_SUPP, data=collegescorecard)
null <- lm(MN_EARN_WNE_P10~1, data=collegescorecard)
step(full, data=collegescorecard, scope=list(lower=null, upper=mod1),
     scale=fullMSE, direction="backward")
```
















