---
title: ''
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readr)
library(mosaic)
collegescorecard <- read_csv("group-h-csv.csv", na = c("", "NA", "NULL", "PrivacySuppressed"),
    col_types = cols(ADM_RATE = col_number(), 
        ADM_RATE_ALL = col_number(), C150_4 = col_number(), 
        CONTROL = col_character(), COSTT4_A = col_number(), 
        FAMINC = col_number(),
        GRAD_DEBT_MDN10YR_SUPP = col_number(), 
        GRAD_DEBT_MDN_SUPP = col_number(), 
        ICLEVEL = col_character(), 
        MD_EARN_WNE_P10 = col_number(), MD_FAMINC = col_number(), 
        MN_EARN_WNE_P10 = col_number(),  
        NPT4_PRIV = col_number(), NPT4_PUB = col_number(), 
        OPEID = col_character(), PCTFLOAN = col_number(), 
        PCTPELL = col_number(),  
        PFTFAC = col_number(), PREDDEG = col_character(), 
        RPY_3YR_RT_SUPP = col_number(), SAT_AVG = col_number(), 
        UNITID = col_character()))
```

#Data set
##Removed the NA values in order to do predictor selection

```{r}
collegescorecard<- collegescorecard%>% 
  select(ADM_RATE, ADM_RATE_ALL, C150_4, CONTROL, COSTT4_A, FAMINC, GRAD_DEBT_MDN10YR_SUPP, GRAD_DEBT_MDN_SUPP, ICLEVEL, MD_EARN_WNE_P10, MD_FAMINC, MN_EARN_WNE_P10, NPT4_PRIV, NPT4_PUB, OPEID, PCTFLOAN, PCTPELL, PFTFAC, SAT_AVG, UNITID) %>%
  mutate( CONTROL=as.factor(CONTROL), ICLEVEL = as.factor(ICLEVEL)) 

nonadata <- na.omit(collegescorecard[c("MN_EARN_WNE_P10","MD_FAMINC","ADM_RATE_ALL","C150_4","SAT_AVG","GRAD_DEBT_MDN_SUPP","COSTT4_A","PFTFAC","PCTPELL","NPT4_PUB","PCTFLOAN", "CONTROL")])

```


#Multicollinearity: we will use COSTT4_A (average annual cost of tuition)

```{r}
nonadata%>%
  select(COSTT4_A, NPT4_PUB)%>%
  cor()

nonadata%>%
  select(PCTPELL, PCTFLOAN)%>%
  cor()
```

#Stepwise Regression
```{r}
library(leaps)
full <- lm(MN_EARN_WNE_P10~MD_FAMINC+
          ADM_RATE_ALL+
             C150_4+
             SAT_AVG
           +GRAD_DEBT_MDN_SUPP
           +COSTT4_A
           +PFTFAC+
             PCTPELL+
             PCTFLOAN, data=nonadata)

null <- lm(MN_EARN_WNE_P10~1, data=nonadata)

fullMSE<-(summary(full)$sigma)^2 


step(null, data=nonadata, scope=list(lower=null, upper=full),
     scale=fullMSE, direction="both")
```

#Final Model
```{r}
stepmod<-lm(formula = MN_EARN_WNE_P10 ~ C150_4 + PCTFLOAN + COSTT4_A + 
    MD_FAMINC + ADM_RATE_ALL + SAT_AVG, data = nonadata)

summary(stepmod)
```

$\widehat{Earnings} = 21,970 +.09144*Median Family Income - 6,441*Admission Rate + 11,400*Completion Rate + 16.66*Average SAT + .5526*Average Cost of Attendance - 16,470*Percent On Federal Loan$


#Checking Conditions

##Good equal variance/linearity, Normality is not met (should we transform?)
##Assuming independence because it's self reported, but schools near each other may have effects?

```{r}
plot(stepmod)
```

##VIF

```{r}
library(car)

vif(full)
vif(stepmod)
```

```{r}
modcor3<-lm(PCTPELL~MD_FAMINC, data=nonadata)
summary(modcor3)
modcor4<-lm(PCTPELL~PCTFLOAN, data=nonadata)
summary(modcor4)
```


#ANOVA

```{r}
anova(stepmod)
```

#Nested F Test: testing subset of percent on pell and percent on a federal loan

```{r}
modelf<-lm(formula = MN_EARN_WNE_P10 ~ MD_FAMINC + ADM_RATE_ALL + C150_4 + 
    SAT_AVG + COSTT4_A + PCTFLOAN + PCTPELL, data = nonadata)
  
modelr<-lm(formula = MN_EARN_WNE_P10 ~ MD_FAMINC + ADM_RATE_ALL + C150_4 + 
    SAT_AVG + COSTT4_A, data = nonadata)

anova(modelr, modelf)
```

#Individual t-tests for variables we thought would be significant but were not in the model

##Percent of full-time faculty
```{r}
pftfacmod<-lm(MN_EARN_WNE_P10~PFTFAC, data=nonadata)
summary(pftfacmod)
```

#Test its effectiveness of the model for some schools in our towns?

#ANOVA test about the different means of the debt repayment groups? If we want to

# Summary statistics

```{r}
favstats(nonadata$MN_EARN_WNE_P10)

favstats(nonadata$MD_FAMINC)

favstats(nonadata$ADM_RATE_ALL)

favstats(nonadata$PCTFLOAN)

favstats(nonadata$COSTT4_A)

favstats(nonadata$SAT_AVG)
```


 












