---
title: ''
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readr)
library(mosaic)
collegescorecard <- read_csv("group-h-csv.csv", na = c("", "NA", "NULL", "PrivacySuppressed"),
    col_types = cols(ADM_RATE = col_number(), 
        ADM_RATE_ALL = col_number(), C150_4 = col_number(), 
        CONTROL = col_character(), COSTT4_A = col_number(), 
        FAMINC = col_number(),
        GRAD_DEBT_MDN10YR_SUPP = col_number(), 
        GRAD_DEBT_MDN_SUPP = col_number(), 
        ICLEVEL = col_character(), 
        MD_EARN_WNE_P10 = col_number(), MD_FAMINC = col_number(), 
        MN_EARN_WNE_P10 = col_number(),  
        NPT4_PRIV = col_number(), NPT4_PUB = col_number(), 
        OPEID = col_character(), PCTFLOAN = col_number(), 
        PCTPELL = col_number(),  
        PFTFAC = col_number(), PREDDEG = col_character(), 
        RPY_3YR_RT_SUPP = col_number(), SAT_AVG = col_number(), 
        UNITID = col_character()))
```

#Data set
##Removed the NA values in order to do predictor selection

```{r}
collegescorecard<- collegescorecard%>% 
  select(ADM_RATE, ADM_RATE_ALL, C150_4, CONTROL, COSTT4_A, FAMINC, GRAD_DEBT_MDN10YR_SUPP, GRAD_DEBT_MDN_SUPP, ICLEVEL, MD_EARN_WNE_P10, MD_FAMINC, MN_EARN_WNE_P10, NPT4_PRIV, NPT4_PUB, OPEID, PCTFLOAN, PCTPELL, PFTFAC, SAT_AVG, UNITID) %>%
  mutate( CONTROL=as.factor(CONTROL), ICLEVEL = as.factor(ICLEVEL)) 

nonadata <- na.omit(collegescorecard[c("MN_EARN_WNE_P10","MD_FAMINC","ADM_RATE_ALL","C150_4","SAT_AVG","GRAD_DEBT_MDN_SUPP","COSTT4_A","PFTFAC","PCTPELL","NPT4_PUB","PCTFLOAN", "CONTROL")])
```


#Multicollinearity: we will use COSTT4_A (average annual cost of tuition)

```{r}
nonadata%>%
  select(COSTT4_A, NPT4_PUB)%>%
  cor()
```

#Backward Elimination
```{r}
library(leaps)
full <- lm(MN_EARN_WNE_P10~MD_FAMINC+
          ADM_RATE_ALL+
             C150_4+
             SAT_AVG
           +GRAD_DEBT_MDN_SUPP
           +COSTT4_A
           +PFTFAC+
             PCTPELL+
             PCTFLOAN, data=nonadata)

null <- lm(MN_EARN_WNE_P10~1, data=nonadata)

fullMSE<-(summary(full)$sigma)^2 

step(full, data=nonadata,
     scale=fullMSE, direction="backward")
```

```{r}
step(full, data=nonadata,
     scale=fullMSE, direction="forward")

```



#Stepwise: gives same model

```{r}
step(null, data=nonadata, scope=list(lower=null, upper=full),
     scale=fullMSE, direction="both")
```



model summary
```{r}
backwardmod<-lm(formula = MN_EARN_WNE_P10 ~ MD_FAMINC + ADM_RATE_ALL + C150_4 + 
    SAT_AVG + COSTT4_A + PCTFLOAN, data = nonadata)

summary(backwardmod)

stepmod<-lm(formula = MN_EARN_WNE_P10 ~ C150_4 + PCTFLOAN + COSTT4_A + 
    MD_FAMINC + ADM_RATE_ALL + SAT_AVG, data = nonadata)

summary(stepmod)
```

$\widehat{Earnings} = 21,970 +.09144*Median Family Income - 6,441*Admition Rate + 11,400*Completion Rate + 16.66*Average SAT + .5526*Average Cost of Attendance - 16,470*Percent On Federal Loan$

# Stepwise with logMD_FAMINC

```{r}
full2<-lm(MN_EARN_WNE_P10~logMD_FAMINC+
          ADM_RATE_ALL+
             C150_4+
             SAT_AVG
           +GRAD_DEBT_MDN_SUPP
           +COSTT4_A
           +PFTFAC+
             PCTPELL+
             PCTFLOAN, data=nonadata)
step(null, data=nonadata, scope=list(lower=null, upper=full2),
     scale=fullMSE, direction="both")
```

```{r}
stepmod2<-lm(MN_EARN_WNE_P10 ~ C150_4 + PCTFLOAN + COSTT4_A + 
    SAT_AVG + ADM_RATE_ALL + logMD_FAMINC, data = nonadata)
summary(stepmod2)
```

```{r}
plot(stepmod2)
```

# Experimenting with higher order models

```{r}
hordmod1<-lm(MN_EARN_WNE_P10 ~ C150_4 + MD_FAMINC*PCTFLOAN + COSTT4_A*MD_FAMINC + C150_4*ADM_RATE_ALL + PCTFLOAN + COSTT4_A + 
    SAT_AVG + ADM_RATE_ALL + MD_FAMINC, data = nonadata)
summary(hordmod1)
```

# Stepwise with interaction terms

```{r}
full3<-lm(MN_EARN_WNE_P10~C150_4*ADM_RATE_ALL+ MD_FAMINC*PCTFLOAN + MD_FAMINC*COSTT4_A + GRAD_DEBT_MDN_SUPP*MD_FAMINC +PCTPELL*PCTFLOAN + PCTPELL*MD_FAMINC + MD_FAMINC+
          ADM_RATE_ALL+
             C150_4+
             SAT_AVG
           +GRAD_DEBT_MDN_SUPP
           +COSTT4_A
           +PFTFAC+
             PCTPELL+
             PCTFLOAN, data=nonadata)
step(null, data=nonadata, scope=list(lower=null, upper=full3),
     scale=fullMSE, direction="both")
```

```{r}
full4<-lm(MN_EARN_WNE_P10~C150_4*ADM_RATE_ALL+ MD_FAMINC*PCTFLOAN + MD_FAMINC*COSTT4_A + GRAD_DEBT_MDN_SUPP*MD_FAMINC + ADM_RATE_ALL*MD_FAMINC + C150_4*MD_FAMINC + SAT_AVG*MD_FAMINC + SAT_AVG*ADM_RATE_ALL +PCTPELL*PCTFLOAN + PCTPELL*MD_FAMINC + MD_FAMINC+
          ADM_RATE_ALL+
             C150_4+
             SAT_AVG
           +GRAD_DEBT_MDN_SUPP
           +COSTT4_A
           +PFTFAC+
             PCTPELL+
             PCTFLOAN, data=nonadata)
step(null, data=nonadata, scope=list(lower=null, upper=full4),
     scale=fullMSE, direction="both")
```

```{r}
interactmod1<-lm(MN_EARN_WNE_P10 ~ C150_4 + PCTFLOAN + COSTT4_A + 
    MD_FAMINC + ADM_RATE_ALL + SAT_AVG + PCTPELL + C150_4:ADM_RATE_ALL + 
    PCTFLOAN:MD_FAMINC + COSTT4_A:MD_FAMINC + PCTFLOAN:PCTPELL, 
    data = nonadata)
summary(interactmod1)
```

```{r}
interactmod2<- lm(MN_EARN_WNE_P10 ~ C150_4 + PCTFLOAN + COSTT4_A + 
    MD_FAMINC + ADM_RATE_ALL + SAT_AVG + PCTPELL + C150_4:MD_FAMINC + 
    C150_4:ADM_RATE_ALL, data = nonadata)
summary(interactmod2)
```




#Low R^2

I don't think the $R^2$ is that low? If we can predict 55% of the variability that's pretty good!



#Checking Conditions

##Good equal variance/linearity, Normality is not met (should we transform?)
##Assuming independence because it's self reported, but schools near each other may have effects?

```{r}
plot(backwardmod)
```

```{r}
plot(stepmod)
```

##VIF

```{r}
library(car)
```

```{r}
vif(full)
vif(stepmod)
```

```{r}
modcor3<-lm(PCTPELL~MD_FAMINC, data=nonadata)
summary(modcor3)
modcor4<-lm(PCTPELL~PCTFLOAN, data=nonadata)
summary(modcor4)
```


#ANOVA

```{r}
anova(backwardmod)
anova(stepmod)
```

#Nested F Test: testing subset of percent on pell and percent on a federal loan

```{r}
modelf<-lm(formula = MN_EARN_WNE_P10 ~ MD_FAMINC + ADM_RATE_ALL + C150_4 + 
    SAT_AVG + COSTT4_A + PCTFLOAN + PCTPELL, data = nonadata)
  
modelr<-lm(formula = MN_EARN_WNE_P10 ~ MD_FAMINC + ADM_RATE_ALL + C150_4 + 
    SAT_AVG + COSTT4_A, data = nonadata)

anova(modelr, modelf)
```

#Individual t-tests for variables we thought would be significant but were not in the model

##Percent of full-time faculty
```{r}
pftfacmod<-lm(MN_EARN_WNE_P10~PFTFAC, data=nonadata)
summary(pftfacmod)
```




#Test its effectiveness of the model for some schools in our towns?

#ANOVA test about the different means of the debt repayment groups? If we want to

# Summary statistics

```{r}
qplot(nonadata$MN_EARN_WNE_P10,
      geom="histogram",  
      main = "Histogram for Mean Earnings", 
      xlab = "Mean Earnings",  
      bins=60,
      fill=I("blue"),
      ylab="Count")

favstats(nonadata$MN_EARN_WNE_P10)
```



```{r}
qplot(nonadata$MD_FAMINC,
      geom="histogram",  
      main = "Histogram for Median Family Income", 
      xlab = "Median Family Income",  
      bins=60,
      fill=I("purple"),
      ylab="Count")

favstats(nonadata$MD_FAMINC)

```

```{r}
nonadata<- nonadata%>% mutate(logMD_FAMINC=log(MD_FAMINC))
qplot(nonadata$logMD_FAMINC,
      geom="histogram",  
      main = "Histogram for Log Median Family Income", 
      xlab = "Log Median Family Income",  
      bins=60,
      fill=I("purple"),
      ylab="Count")

favstats(nonadata$logMD_FAMINC)
```

```{r}
nonadata<- nonadata%>% mutate(sqrtMD_FAMINC2=sqrt(MD_FAMINC))
qplot(nonadata$MD_FAMINC2,
      geom="histogram",  
      main = "Histogram for SQRT Median Family Income", 
      xlab = "SQRT Median Family Income",  
      bins=60,
      fill=I("purple"),
      ylab="Count")

favstats(nonadata$sqrtMD_FAMINC)
```

```{r}
qplot(nonadata$ADM_RATE_ALL,
      geom="histogram",  
      main = "Histogram for Admission Rate", 
      xlab = "Admission Rate",  
      bins=60,
      fill=I("red"),
      ylab="Count")

favstats(nonadata$ADM_RATE_ALL)
```

```{r}
qplot(nonadata$PCTFLOAN,
      geom="histogram",  
      main = "Histogram for Percent on Federal Loan", 
      xlab = "Percent on Federal Loan",  
      bins=60,
      fill=I("orange"),
      ylab="Count")

favstats(nonadata$PCTFLOAN)
```


```{r}
qplot(nonadata$COSTT4_A,
      geom="histogram",  
      main = "Histogram for Average Cost of Attendance", 
      xlab = "Average Cost of Attendance",  
      bins=60,
      fill=I("green"),
      ylab="Count")

favstats(nonadata$COSTT4_A)
```


```{r}
qplot(nonadata$SAT_AVG,
      geom="histogram",  
      main = "Histogram for Average SAT Score", 
      xlab = "Average SAT Score",  
      bins=60,
      fill=I("yellow"),
      ylab="Count")

favstats(nonadata$SAT_AVG)
```












